\myTitle

### 2015-03-12 tex/response_to_refs_letter.tex

\end{itemize}

{\bf Referee Number 1 comments}

I have several important concerns over this paper:

\reply{I'll give brief replies to each of the following here in outline,
and will give more complete details in the following sections.}
\begin{enumerate}
\item The conceptual framework is inappropriate since FLOCK is not an MCMC algorithm
\reply{{\sc flock} is not an MCMC algorithm but it {\em is} an algorithm and an inference procedure and 
recognizing it as a special case of well known statistical methods is appropriate.}
\item There is only one dataset which is way too little to come to conclusions
\reply{We agree with the referee. We have undertaken a simulation study across a range of different $F_\mathrm{ST}$ values.}
\item There are basic flaws in the interpretation of the outputs of FLOCK, leading to invalid
comparisons relative to K values other than the estimate
\reply{We have eliminated issues of what can and can't be said about allocation of individuals  at a
certain value of $K$ by doing all the simulations and the analyses at a known value of $K$.}
\item Comparisons of estimates of K cannot be performed with sufficient rigor
\reply{We agree with the referee here.  In fact, we weren't entirely comfortably addressing the issue of estimating
$K$ in the previous version of the manuscript, lest it appear that we were implicitly encouraging it. 
Therefore, we now focus entirely on allocation of individuals.}
\item The speed comparisons are based on an ill-chosen processing unit
\reply{We implemented the {\sc flock} algorithm by modifying the C source code for {\sc structure} and
adding an R wrapper.  This allows a more direct comparison of the amount of computation involved.  If 
{\sc flock} were implemented as efficiently as {\sc structure} it would be much faster.  But since it
runs 200 times slower than it should for what it is doing, there doesn't seem much speed advantage in 
the data sets we ran.  We report that now in the paper in a way that should be easily understood. }
\end{enumerate}

@@

The current popularity of MCMC methods in the field of genetic analysis does not justify trying to cast every current algorithm into an MCMC framework.

\reply{Again, we are not so much casting it in an MCMC framework as we are in a 
statistical/probabilistic framework.  Since it is a problem in statistical inference
it seems to us reasonable to understand it in terms of the probabilities involved. Those
probabilities are, for the most part, identical to those in the program {\sc structure} because
the underlying model is the same in the two programs.  Regarding casting the algorithm in
comparison to {\sc structure}, it makes sense to us to try to understand and explain the {\sc flock}
algorithm in those terms since {\sc structure} is so well known and because the models
underlying it and {\sc flock} are essentially identical.  }

For example:
\begin{itemize}
\item Line 156: it is suggested that Flock can be trapped in a local mode, this is incorrect because Flock does NOT search for maximum values, either local or else
\reply{We thank the referee for pointing out our erroneous language here.  We should not have used
the word ``trapped.''  We have updated that sentence to read, ``As such, we should not be surprised
if it is prone 
to converging to different parts of the posterior probability surface on runs started
from different initial conditions.''}

\item Line 254: it is suggested that Flock is searching over all possible partitions. Flock does not sample nor search the surface or space of all partitions.
\reply{We would not have wanted it to sound like we thought {\sc flock} was searching
each and every possible
partition, since this would be infeasible in most problems, so we thank the referee for pointing out that our reference to a space of all partitions might
have confused readers in this way.  We have changed that section appropriately.  However, on whether
or not {\sc flock} samples from the space of partitions, we respectfully submit that the referee is
incorrect when he states that, ``Flock does not sample nor search the surface or space of all partitions.'' 
Consider this: 1) for each run, the initial starting partition in {\sc flock} is chosen (or you could say,
sampled) randomly from amongst
possible partitions with roughly the same number of individuals in each cluster; 2) the starting partition uniquely
determines the partition that the algorithm will converge to since it is deterministic, 3) different
starting partitions can lead to different final solutions/partitions.  How is it then that each
run in {\sc flock} is {\em not} a sample from the partitions?  }
\item Line 234: ``individuals would alternate between reference groups'': this, precisely, is an orbit
\reply{We removed discussions of individuals alternating between reference groups.}
\end{itemize}

@@

{\bf 2) Data}

Performance comparisons between programs based on a single dataset are of little value. The fact that it is large and real does not compensate. For this purpose, one would expect that a large set of simulations spanning several parameter values with or without several real datasets would be used instead of a single empirical dataset. The advantage of simulations is, of course, that the true genetic structure of the data (essentially K) is known and so the relative performances may be compared on a solid basis. Here, unless I missed something, there is no independent (of the two programs) source of knowing with a reasonable degree of certainty the actual number of clusters represented within this collection of genotypes. In fact, there appears to be several candidate values for K with this particular dataset (see below).

\reply{We agree whole-heartedly and undertake numerous simulations of data like the real
steelhead data in the previous submission.  Having done this, and analyzed the results,
we can now address the comments below.}
Several (over)generalizations are based on this unique dataset:
\begin{itemize}
\item (p.9 ) FLOCK seems to find the same solution less consistently than does STRUCTURE
\reply{From our simulations using {\sc flockture} this is clearly true.  One need only look
at the variability present in the purple boxplots in Figures~2 and~3 to see that there is more
variability between {\sc flockture} analyses (50 runs of 20 iterations each) on the same data set 
than between {\sc structure} analyses. }
\item (p.14) The problem of getting caught in local modes is not unique to FLOCK...; however our results suggest it is a bigger problem for FLOCK
\reply{The simulation results bear this out, but we have reworded it to, ``This likely 
increases the number of
local peaks, leading {\sc flockture} to settle on different solutions in 
different runs. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for {\sc flockture}.''
}
\item (p.16) Given FLOCK'S tendency to converge to different solutions with large K and n...
\reply{See above.  Note that we don't vary $K$ now so we don't call it a problem ``with large $K$'' any
longer.}
\item (p.16) It is probably better to use STRUCTURE in such cases...
\reply{We don't use this phrase specifically, but we do point out that {\sc flock} is not best-suited to high-gene-flow
species with very low $F_\mathrm{ST}$.}
\end{itemize}

@@

{\bf 3) Interpretation of FLOCK outputs (purpose of FLOCK, best run, stochasticity of plateau lengths)
Purpose of FLOCK}

I agree with the following statement (p. 11):
``While we take the stance that estimates of K made with any unsupervised clustering algorithm from data on real (non-idealized) populations should always be interpreted and used cautiously...''
I am fully aware that we are dealing with probabilistic statements here, not certainties. However, I am also convinced that the search for the number of genetic units (K) is a central issue. This is precisely what Flock intends to do, and it is also the main task that STRUCTURE was built for.

\reply{We disagree with the referee that estimating $K$ was the ``main task that {\sc structure} was
built for.''  I had the pleasure of visiting Jonathan Pritchard in Oxford shortly after the
{\sc structure} paper was published. At that time, the members of the Donnelly lab were not eagerly using
{\sc structure} to estimate $K$ from different samples; they were using {\sc structure} to correct
for population admixture when doing association tests in human genetics---our understanding
is {\em that} was the task that structure was original built for.  Furthermore, when I contacted
Jonathan a few years later to ask him some details about how $\ln P(D)$ was calculated by {\sc structure},
he pointed out to me (while getting his notes out of his filing cabinet) that the method for estimating
$K$ was added into the paper to satisfy a referee who had requested it, and that Matthew Stephens 
(one of the co-authors) was not ever extremely pleased to have it in there.  Finally, I teach an annual
course with Matthew Stephens, and each year he explains to the students why estimating $K$ is so
difficult and why estimates of $K$ from {\sc structure} or any other procedure should be 
regarded with circumspection.  }

### 2015-03-12 tex/main-body-text.tex
Further tweaks and edits.  Very close to resubmission.

Given the above sort of emphasis on the distinctiveness of the {\sc flock}
algorithm, it would not be surprising if most readers incorrectly believed {\sc flock}
to be based upon a completely different model than the well-known program
{\sc structure}. Here we remedy such confusion with a simple derivation
that identifies 
the close relationship between the {\sc flock} algorithm and the program
{\sc structure}.  Although the algorithm in {\sc flock}
is, as the authors point out, deterministic (apart from a random initial allocation of 
individuals to clusters), it can be interpreted as a 
limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
models.  More specifically, the {\sc flock} 
algorithm is a special case of the simulated annealing
algorithm for finding the Bayesian maximum-{\em a-posteriori}
estimate from a marginalized form of the {\sc structure} model with no
admixture and non-correlated allele frequencies.  Thus, while the models underlying
{\sc structure} and {\sc flock} are nearly indistinguishable, {\sc structure}'s inference
relies on MCMC samples from the posterior distribution, while
{\sc flock}'s inference involves finding a local maximum in the posterior surface.

@@

\section*{Comparison of {\sc structure} and {\sc flock}}
{\sc flock}, initially designed to map admixture in hybrid populations 
\citep{Duc&Tur2009}, was recently proposed as an alternative to 
{\sc structure} for estimating the number of populations from a collection with 
unsampled source populations \citep{Duc&Tur2012}. Whatever the ultimate goal
of its application,  
{\sc flock} relies on clustering individuals with like genotypes together  
through an iterative process. Thus, a reasonable way 
to gauge how well the algorithm functions is to simulate data of known 
population structure with varying levels of differentiation and 
evaluate how well it does in recovering that structure when the data are
analyzed assuming
the number of subpopulations is exactly the number of subpopulations in the simulation. The 
algorithm performance can then be measured with a zero-one 
loss function (where a loss of 1 is incurred if an individual is group with the wrong cluster).

The implementation of {\sc flock} as a VBA procedure (macro) in 
Microsoft Excel retards rigorous evaluation of its behavior due 
the its slow execution speed and 
the inability to efficiently batch-process multiple data sets. Consequenty, we adapted 
the source code from {\sc structure} as described above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) 
in order to accurately assess the strengths and weaknesses of the {\sc flock} 
algorithm. First, using simulated data we demonstrate that {\sc flock} 
and {\sc flockture} are essentially equivalent programs.
Second, we simulate and analyze datasets with {\sc flockture} 
and three models in {\sc structure} (the admixture and correlated 
allele frequency (A.C) model, the no-admixture and correlated 
allele frequency (NA.C) model, and the no-admixture and non-correlated 
allele frequency (NA.NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure}.

For each set of runs, {\sc flock} reports the 
normalized likelihood values (the exponentiated log-likelihood value for an individual's 
membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the 
longest plateau---refered to as the `best run' by \citep{Duc&Tur2012} (see below).
These values can be compared to the $q_i$ values
returned by {\sc structure}, which, in the no-admixture case, are posterior 
probabilities of cluster membership. In {\sc flock} the user must specify both the number of runs 
(different initial allocations of individuals 
to each cluster) as well as the number of iterations (in each iteration every individual 
has the opportunity to be reallocated 
to a different cluster).  At the conclusion of each run, the log-likelihood value for 
membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the 
cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} 
identifies runs that have converged to the same solution by detecting 
that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same 
solution are termed a ``plateau.'' The plateau record is a list of 
how many runs converged to the same allocation
of individuals.

To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
were simulated using a coalescent based program (code provided on GitHub repository 
https://github.com/eriqande/flock-comment/simdata). The 5 populations
were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed  at 15 loci.
Each of the four data sets was simulated with a different 
average pairwise $F_{ST}$ between the populations: 0.03, 0.02, 0.01, and 0.005. 
Both {\sc flock} and {\sc flockture} were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate 
representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into corrected population as
 defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
RAM running 64-bit Windows 7.

Once we were confident that no major differences in inference existed between 
{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for each of two marker types 
(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader 
range of population differentiation. The 96 datasets were simulated as above (five populations 
composed of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes for the microsatellites
and 96 loci for the SNPs)
with average pairwise $F_{ST}$ values ranging between 0.004 and 0.045. 
 {\sc flockture} was run with the default run parameters and each run of 
{\sc structure} consisted of a 5,000 sweep burn-in followed by a 20,000 sweep sample
from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
was used to relabel clusters among different runs,
while the program {\sc distruct} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

@@

We attempted for some time to develop {\sc flockture} to yield
completely identical results to those from {\sc flock}; however we did not
persist in those endeavors when we realized that {\sc flock} itself does
not provide consistent results when expected by symmetry.  Briefly, In an attempt to understand the 
we took a subset of the 
first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
{\sc flock} with all individuals originating from population 1 being placed in reference population 1 and all 
individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
with the same starting clusters, but with the reference population labels switched. While the final allocation
of individuals matched, the mean LLOD was slightly different (5.045009 vs. 5.028084), indicating it would
be futile to write {\sc flockture} to yield formally identical results to {\sc flock}---{\sc flock} itself
doesn't yield identical results when it should. Therefore, comparisons
between {\sc flock} and {\sc flockture} were limited to random initial starting points and so will not 
be identical, but do display equivalent behavior.

@@

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Infrequently, at higher levels of population differentiation the 
NA.NC and NA.C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
resulting from the assignment of individuals from two groups into a single cluster while 
individuals from another group were separated into two clusters
with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct} A). This phenomenon 
(only present in a single microsatellite dataset) is an instance of {\sc structure}'s
Markov chain failing to converge well in the short runs we were using. As the subtlety of genetic structure 
increased loss for both programs increased and differences in the performance of
each model with different marker types emerged.
For the microsatellite dataset we observed that both programs
produce roughly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increased and certain models did marginally
better in assignment. No model did consistently better than others, until $F_{ST}$ dropped 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
{\sc flockture} did, however, have consistently smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016, however, {\sc structure} began correctly assigning a 
higher proportion of individuals.

The number and length of plateaus {\sc flockture} encountered quickly 
diminished as the amount of differentiation among populations decreased. For the microsatellite datasets,
when $F_{ST}$ was less than 0.0337, no plateau record
contained a plateau of length six or greater. No plateaus were observed 
(i.e. no two runs of {\sc flockture} gave exactly the same solution) when 
$F_{ST}$ was smaller than 0.02187. Results from the SNP datasets
were concordant.
Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets, 
graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
likelihood were suggestive of population structure.

@@

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
  \caption{
  Program {\sc distruct} plots of cluster membership probabilities of simulated individuals
  ($q_i$ values from {\sc structure} and the normalized likelihood values from {\sc flock}). 
  Individuals are grouped left to right into their simulated population samples, and each 
  horizontal plot represents a run of the program
  {\sc structure}, or the ``longest-plateau-run'' from a series of runs in {\sc flock} (unless
  there were no plateaus, in which case the run reaching the highest likelihood was chosen).
  Within horizontal plots, each individual is represented by a vertical bar composed of colored
  segments of lengths proportional to the various cluster membership probabilities.
  Each horizontal plot
  represents one of nine 
 runs of the programs  at $K$ = 5 (the true value of $K$). Three models were run in structure: the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model. Numbers in 
parentheses following the model names give the number, out of 9, of replicate runs
that obtained the same visually-indistinguishable estimate. Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s `best-run' solutions consistently assigned fewer individuals correctly
and became inconsistent at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence which
makes it possible for many different allocations of individuals to
have roughly the same likelihood.
The likelihood surface defined on partitions of this space is then often
be multimodal---possessing numerous local peaks and troughs.
The peak that  {\sc flockture}'s algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} to settle on many different solutions in each run. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for {\sc flockture}.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each execution of {\sc flock}.
\citet{Duc&Tur2012} observed faster processing times relative to {\sc structure}.
Having implemented {\sc flockture} it is clear that the amount of computation required for a
single sweep of {\sc structure} is equivalent to an iteration of reallocation of individuals
in {\sc flock}.
Hence, if {\sc flock} and {\sc structure} took similar amounts of time for comparable
computations, one would expect that running flock once with 50 runs, each of 20 iterations,
would require about the same amount of time as running {\sc structure} on the same
data set for 1,000 sweeps.  Since it is recommended to run {\sc structure} longer than 1,000 sweeps, {\sc flock} should have a speed advantage. However, {\sc flock} requires far more 
time for each computation
than {\sc structure}, as is clear from the fact that it runs 200 times slower than 
{\sc flockture}.  Thus, for data sets comparable to the ones we simulated, one can expect
50 runs of 20 iterations in {\sc flock} (a standard execution of the program) to require as
much time as running {\sc structure} for 200,000 sweeps. This does not constitute
a speed advantage.
If {\sc flock} were to be implemented using good coding practices in a fast language,
rather than as a VBA routine in Microsoft Excel,
it could be made much faster. As it is, we found it preferable to implement our own version
in {\sc flockture} for our simulation assessment of the {\sc flock} algorithm.

Many researchers wish to estimate $K$, the number of subpopulations, from genetic data.  This is
a difficult task, statistically, because estimating $K$ is a problem in model selection, and hence can
be exquisitely sensitive to the priors on nuisance parameters as well as to departures from the underlying
model (e.g. clinal rather than demic structure). For these reasons, it seems inadvisable to us to
put too much stock in the inference of $K$ from any program with a simple model of population structure which
is almost undoubtedly violated in reality. Accordingly, we have based our comparison of 
{\sc flockture} and {\sc structure} on their ability to correctly cluster individuals when the true value
of $K$ is known. It should always be kept in mind that real data may depart in unexpected ways from
simulated data.

>>>>

Given the above sort of emphasis on the distinctiveness of the {\sc flock}
algorithm, it would not be surprising if most readers incorrectly believed {\sc flock}
to be based upon a completely different model than the well-known program
{\sc structure}. Here we remedy such confusion with a simple derivation
that identifies 
the close relationship between the {\sc flock} algorithm and the program
{\sc structure}.  Although the algorithm in {\sc flock}
is, as the authors point out, deterministic (apart from a random initial allocation of 
individuals to clusters), it can be interpreted as a 
limiting case of a probabilistic search algorithm applied to one of {\sc structure}'s
models.  More specifically, the {\sc flock} 
algorithm is a special case of the simulated annealing
algorithm for finding the Bayesian maximum-{\em a-posteriori}
estimate from a marginalized form of the {\sc structure} model with no
admixture and non-correlated allele frequencies.  Thus, while the models underlying
{\sc structure} and {\sc flock} are nearly indistinguishable, {\sc structure}'s inference
relies on MCMC samples from the posterior distribution, while
{\sc flock}'s inference involves finding local maxima in the posterior surface.

@@

\section*{Comparison of {\sc structure} and {\sc flock}}
{\sc flock}, initially designed to map admixture in hybrid populations 
\citep{Duc&Tur2009}, was recently proposed as an alternative to 
{\sc structure} for estimating the number of populations from a collection with 
unsampled source populations \citep{Duc&Tur2012}. Whatever the ultimate goal
of its application,  
{\sc flock} relies on clustering individuals with like genotypes together  
through an iterative process. Thus, a reasonable way 
to gauge how well the algorithm functions is to simulate data of known 
population structure with varying levels of differentiation and 
evaluate how well it does in recovering that structure when the data are
analyzed assuming
the number of subpopulations is exactly the number of subpopulations in the simulation. The 
algorithm performance can then be measured with a zero-one 
loss function (where a loss of 1 is incurred for each individual that is grouped with the wrong cluster).

The implementation of {\sc flock} as a VBA procedure (macro) in 
Microsoft Excel retards rigorous evaluation of its behavior due 
to its slow execution speed and 
the inability to efficiently batch-process multiple data sets. Consequenty, we adapted 
the source code from {\sc structure} as described above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) 
in order to accurately and reproducibly assess the strengths and weaknesses of the {\sc flock} 
algorithm. First, using simulated data we demonstrate that {\sc flock} 
and {\sc flockture} are essentially equivalent programs.
Second, we simulate and analyze datasets with {\sc flockture} 
and three models in {\sc structure} (the admixture and correlated 
allele frequency (A.C) model, the no-admixture and correlated 
allele frequency (NA.C) model, and the no-admixture and non-correlated 
allele frequency (NA.NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure}.

For a single excecution of {\sc flock} the user must specify both the number of runs 
(different initial allocations of individuals 
to each cluster) as well as the number of iterations (in each iteration every individual 
has the opportunity to be reallocated 
to a different cluster). At the conclusion of each run, the log-likelihood value for 
membership to all $K$ clusters is computed
and the log likelihood difference (LLOD)---the difference between the 
cluster with the highest likelihood 
and second highest likelihood---is calculated for each individual. {\sc flock} 
identifies runs that have converged to the same solution by detecting 
that they have the same mean LLOD score, and, in 
{\sc flock} parlance multiple runs that have converged to the same 
allocation of individuals are termed a ``plateau.'' The plateau record is a list of 
how many runs in the program execution have converged to the same allocation
of individuals.
When all 50  runs have completed, 
{\sc flock} reports as its solution the cluster membership and 
normalized likelihood values (the exponentiated log-likelihood value for an individual's 
membership to each of $K$ clusters, scaled to sum to one) for one of the 50 runs belonging to the 
longest plateau---refered to as the `best run' by \citep{Duc&Tur2012}.
These normalized likelihood values can be compared to the $q_i$ values
returned by {\sc structure}, which, in the no-admixture case, are posterior 
probabilities of cluster membership.

To compare {\sc flock} and {\sc flockture}, 4 datasets composed of 5 populations
were simulated using a coalescent-based program (code provided at 
https://github.com/eriqande/flock-comment/). The 5 populations
were composed of 250, 250, 400, 550, and 550 individuals, respectively, all typed  at 15 loci.
Each of the four data sets was simulated with a different 
average pairwise $F_{ST}$ between the populations: 0.03, 0.02, 0.01, and 0.005. 
Both {\sc flock} and {\sc flockture} were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate 
representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into correct population as
 defined by the zero-one loss function, (2) individual $q_i$ values, and (3) plateau records. 
All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB of 
RAM running 64-bit Windows 7.

Once we were confident that no major differences in inference existed between 
{\sc flock} and {\sc flockture} we used {\sc flockture} to analyze 48 datasets for each of two marker types 
(microsatellites and single nucleotide polymorphisms (SNPs))  that encompassed a broader range
of population differentiation. The 96 datasets were simulated as above (five populations 
composed of 250, 250, 400, 550, and 550 individuals with 15 loci for the microsatellites
and 96 loci for the SNPs)
with average pairwise $F_{ST}$ values ranging between 0.004 and 0.045. 
 {\sc flockture} was run with the default run parameters and each run of 
{\sc structure} consisted of a 5,000 sweep burn-in followed by a 20,000 sweep sample
from the posterior distribution. Each dataset was analyzed 9 times by each of the different
methods. All runs were performed on a Macbook with a 2.3 GHz i7 processor
and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
was used to relabel clusters among different runs,
while the program {\sc distruct} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

When {\sc flock} does not have a `best run' because the plateau record contains no plateaus, the program still returns a solution, but it is not clear how that solution is chosen.  In
our analyses  with {\sc flockture}, when the plateau record did not indicate a `best run,' we used the
run with the highest likelihood as the solution.

@@

We attempted for some time to develop {\sc flockture} to yield
completely identical results to those from {\sc flock}; however we did not
persist in those endeavors when we realized that {\sc flock} itself does
not provide consistent results when expected by symmetry.  Briefly, In an attempt to understand the
inner workings of {\sc flock} 
we took a subset of the 
first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
{\sc flock} with all individuals originating from population 1 being placed in reference population 1 and all 
individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
with the same starting clusters, but with the reference population labels switched. While the final allocation
of individuals matched, the mean LLOD was slightly different (5.045009 vs. 5.028084), indicating it would
be futile to write {\sc flockture} to yield formally identical results to {\sc flock}---{\sc flock} itself
doesn't yield identical results when it should. Therefore, comparisons
between {\sc flock} and {\sc flockture} were limited to random initial starting points and so will not 
be identical, but do display equivalent behavior.

@@

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well-differentiated populations. Infrequently, at higher levels of population differentiation the 
NA.NC and NA.C models in {\sc structure} exhibited large losses (Fig.~\ref{fig:uSatLoss})
resulting from the assignment of individuals from two groups into a single cluster while 
individuals from another group were separated into two clusters
with intermediate $q_i$ values (Fig.~\ref{fig:FvSdistruct} A). This phenomenon 
(only present in a single microsatellite dataset) is an instance of {\sc structure}'s
Markov chain failing to converge well in the short runs we were using. As the subtlety of genetic structure 
increased, loss for both programs increased and differences in the performance of
each model with different marker types emerged.
For the microsatellite dataset we observed that both programs
produce roughly equivalent results at higher levels of population differentiation. At intermediate 
values, the variance of the loss incurred by each model increased and certain models did marginally
better in assignment. No model did consistently better than others, until $F_{ST}$ dropped 
below 0.0136 after which the loss for {\sc flockture} was markedly larger as was the among-run
variance. The drastic increase in variance among runs was also observed with the SNP datasets. 
{\sc flockture} did, however, have consistently smaller loss than  {\sc structure} at high to intermediate 
values of $F_{ST}$. Once $F_{ST}$ fell below 0.016, however, {\sc structure} began correctly assigning a 
higher proportion of individuals.

The number and length of plateaus {\sc flockture} encountered quickly 
diminished as the amount of differentiation among populations decreased. For the microsatellite datasets,
when $F_{ST}$ was less than 0.0337, no plateau record
contained a plateau of length six or greater. No plateaus were observed 
(i.e. no two runs of {\sc flockture} gave exactly the same solution) when 
$F_{ST}$ was smaller than 0.02187. Results from the SNP datasets
were concordant.
Only simulated datasets with pairwise $F_{ST}$ values greater 0.0323 had plateau records with a 
plateau greater than six. Despite the lack of plateaus observed for many of the simulated datasets, 
graphs of the normalized likelihood values from the {\sc flockture} run with the highest 
likelihood were indicative of population structure (Fig~\ref{fig:FvSdistruct}).

@@

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}%B+W version use 'FlockturevStructureDistruct_BW.jpg'
  \caption{
  Program {\sc distruct} plots of cluster membership probabilities of simulated individuals
  ($q_i$ values from {\sc structure} and the normalized likelihood values from {\sc flock}). 
  Individuals are grouped left to right into their simulated population samples, and each 
  horizontal plot represents a run of the program
  {\sc structure}, or the ``longest-plateau-run'' from a series of runs in {\sc flock} (unless
  there were no plateaus, in which case the run reaching the highest likelihood was chosen).
  Within horizontal plots, each individual is represented by a vertical bar composed of colored
  segments of lengths proportional to the various cluster membership probabilities.
  Each horizontal plot
  represents one of nine 
 runs of the programs  at $K$ = 5 (the true value of $K$). Three models were run in structure: the admixture and correlated allele 
  frequency (A.C) model, 
the no-admixture and correlated allele frequency (NA.C) model, 
and the no-admixture and non-correlated allele frequency (NA.NC) model. Numbers in 
parentheses following the model names give the number, out of 9, of replicate runs
that obtained the same visually-indistinguishable estimate. (Note that although many of {\sc flocktures}'s solutions in B--D were visually indistinguishable, they were not identical partitions, and would
hence be considered different for the purposes of creating a {\sc flock} plateau record.) Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

@@

Though we observed similar results
between the two programs at large values of $F_{ST}$, 
{\sc flockture}'s solutions consistently assigned fewer individuals correctly---and the
replicate solutions for individual data sets became more variable---at lower values of $F_{ST}$.
This is likely a consequence of the low population divergence which
makes it possible for many different allocations of individuals to
have roughly the same likelihood.
The likelihood surface defined on partitions of this space is then often
be multimodal---possessing numerous local peaks and troughs.
The peak that the {\sc flock} algorithm finds is entirely determined by the initial, random, allocation of 
individuals, as the algorithm proceeds deterministically after the initial random allocation.
Decreasing $F_{ST}$ leads to more homogenous 
populations whose multilocus genotypes have a smaller difference in 
likelihood of assigning to any given cluster. This likely increases the number of
local peaks, leading {\sc flockture} and {\sc flock} to settle on many different solutions in each run. The problem presented by having multiple modes in the likelihood surface 
is not unique to {\sc flock}---{\sc structure} also can converge to 
different solutions in complex spaces; however our simulations suggest this is a more  
prevalent problem for {\sc flockture}.
{\sc flock} is not best-suited to species characterized by large
subpopulations and low divergence, as is the case with many 
marine fish.

One of the advantages of {\sc flock} over other methods for genetic clustering
is reported to be the short processing time for each execution of {\sc flock}.
\citet{Duc&Tur2012} observed faster processing times relative to {\sc structure}.
Having implemented {\sc flockture} it is clear that the amount of computation required for a
single sweep of {\sc structure} is equivalent to an iteration of reallocation of individuals
in {\sc flock}.
Hence, if {\sc flock} and {\sc structure} took similar amounts of time for comparable
computations, one would expect that executing {\sc flock} once with 50 runs, each of 20 iterations,
would require about the same amount of time as running {\sc structure} on the same
data set for 1,000 sweeps.  Since it is recommended to run {\sc structure} longer than 1,000 sweeps, {\sc flock} should have a speed advantage. However, {\sc flock} requires far more 
time for each computation
than {\sc structure}, as is clear from the fact that it runs 200 times slower than 
{\sc flockture}.  Thus, for data sets comparable to the ones we simulated, one can expect
50 runs of 20 iterations in {\sc flock} (a standard execution of the program) to require as
much time as running {\sc structure} for 200,000 sweeps. This does not constitute
a speed advantage.
If {\sc flock} were to be implemented using good coding practices in a fast language,
rather than as a VBA routine in Microsoft Excel,
it could be made much faster. As it is, we found it preferable to implement our own version
of the {\sc flock} algorithm
in {\sc flockture} for our simulation assessment of the {\sc flock} algorithm.

Many researchers wish to estimate $K$, the number of subpopulations, from genetic data.  This is
a difficult task, statistically, because estimating $K$ is a problem in model selection, and hence can
be exquisitely sensitive to the priors on nuisance parameters as well as to departures from the underlying
model (e.g. clinal rather than demic structure). For these reasons, it seems inadvisable to us to
put too much stock in the inference of $K$ from any program, especially when population structure
is subtle. Accordingly, we have based our comparison of 
{\sc flockture} and {\sc structure} on their ability to correctly cluster individuals when the true value
of $K$ is known.

While there has been lengthy 
discourse over the choice of estimators for $K$ in {\sc structure} 
\citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011}, less has been written on how the number of 
individuals
and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules for $K$ based on simulated data 
sets with a modest number of individuals and different migration regimes and rates, numbers of loci, and $K$. 
By relying on the number of identical partitions obtained by {\sc flock} over multiple
runs as a means of supporting 
a particular $K$,  similar---but not identical---partitions
will provide no support to a given $K$. This could be problematic, 
since we have shown that even at intermediate
values of $F_{ST}$,  {\sc flockture} (and, by extension,
{\sc flock}) may converge to 
different cluster allocations in every run, so that no plateaus are encountered.
